# Reformatting the College Scorecard data

```{r setup, include = FALSE}
library(dplyr)
library(ggplot2)
library(ggformula)
```

The Scorecard data are available at [this website](https://collegescorecard.ed.gov/data/). Documentation [here](https://collegescorecard.ed.gov/data/documentation/).

For the present, I'm using the data I downloaded when originally writing the activity.

```r
download.file("http://tiny.cc/dcf/ScorecardSmall.Rda", 
              destfile = "ScorecardSmall.Rda")
```

The subset includes all 7804 institutions in the original 2013 Scorecard file, but just 54 variables.

The purpose of this document is to reform the data with nicer variable names and a better naming convention for the financial variables that I care about.

```{r}
load("ScorecardSmall.Rda")
names(ScorecardSmall)
table(ScorecardSmall$CONTROL)
```

Renaming variables:
```{r}
Scorecard_small <- ScorecardSmall %>%
  rename(carnegie = CCSIZSET,
         net_tuit_fte = TUITFTE,
         control = CONTROL,
         fac_sal = AVGFACSAL,
         tuition_in = TUITIONFEE_IN,
         tuition_out = TUITIONFEE_OUT,
         rel_affil = RELAFFIL,
         name = INSTNM,
         id = OPEID,
         adm_rate = ADM_RATE,
         low_income = INC_PCT_LO,
         parent_hs = PAR_ED_PCT_HS,
         parent_ms = PAR_ED_PCT_MS,
         parent_post = PAR_ED_PCT_PS,
         first_gen = PAR_ED_PCT_1STGEN,
         median_debt = GRAD_DEBT_MDN_SUPP,
         spend_fte = INEXPFTE) %>%
  mutate(control = c("public", "private", "other")[control]) %>%
  mutate(net_price = ifelse(is.na(NPT4_PUB), NPT4_PRIV, NPT4_PUB)) %>%
  mutate(net_price_q1 = ifelse(is.na(NPT41_PUB), NPT41_PRIV, NPT41_PUB)) %>%
  mutate(net_price_q2 = ifelse(is.na(NPT42_PUB), NPT42_PRIV, NPT42_PUB)) %>%
  mutate(net_price_q3 = ifelse(is.na(NPT43_PUB), NPT43_PRIV, NPT43_PUB)) %>%
  mutate(net_price_q4 = ifelse(is.na(NPT44_PUB), NPT44_PRIV, NPT44_PUB)) %>%
  mutate(net_price_q5 = ifelse(is.na(NPT45_PUB), NPT45_PRIV, NPT45_PUB)) %>%
  mutate(net_price_3075 = mean(na.rm = TRUE, c(NPT4_3075_PRIV, NPT4_3075_PUB, NPT4_3075_OTHER))) %>%
  mutate(net_price_75UP = round(mean(na.rm = TRUE, c(NPT4_75UP_PRIV, NPT4_75UP_PUB, NPT4_75UP_OTHER)))) %>%
  mutate(net_price_048 = round(mean(na.rm = TRUE, c(NPT4_048_PRIV, NPT4_048_PUB, NPT4_048_OTHER)))) %>%
  mutate(ret_ft = round(mean(na.rm = TRUE, c(RET_FT4, RET_FTL4)))) %>%
  mutate(ret_ft = round(mean(na.rm = TRUE, c(RET_PT4, RET_PTL4)))) %>%
  mutate(  adm_rate = round(100 * as.numeric(adm_rate))
         , first_gen = round(100 * as.numeric(first_gen))
         , parent_ms = round(100 * as.numeric(parent_ms))
         , parent_hs = round(100 * as.numeric(parent_hs))
         , parent_post = round(100 * as.numeric(parent_post))
         , low_income = round(100 * as.numeric(low_income))
         , median_debt = as.numeric(median_debt)
         ) %>%
  
  select(-starts_with("NPT4"), -starts_with("RET_"), -CCUGPROF)
  
names(Scorecard_small)
# save(Scorecard_small, file = "../data/Scorecard_small.rda")
```

## Merging the public and private data

```{r}
ScorecardSmall %>%
  group_by(is.na(RET_PT4), is.na(RET_PTL4)) %>%
  tally()
ScorecardSmall %>%
  filter(is.na(NPT4_PUB), is.na(NPT4_PRIV))  -> foo
nrow(foo)
```


## Auxilliary Data

* The `Carnegie.csv` file details the 18 [Carnegie size classifications](http://carnegieclassifications.iu.edu/classification_descriptions/size_setting.php).

