---
title: "The College Scorecard (part I)"
output: rtutor::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(rtutor)
library(tidyverse)
library(ggformula)
library(DT)
library(USCOTS2017DC)
knitr::opts_chunk$set(echo = FALSE)
```

The [College Scorecard](https://collegescorecard.ed.gov/data/) involves data collected by the US Department of Education and the Department of the Treasury about costs and other financial/demographic aspects of higher education on an institution-by-institution basis. The data have hundreds of variables. In this exercise, you'll use an abbreviated form of the data with just `r length(Scorecard_small)` variables.

The table shows a few randomly selected cases from the `Scorecard_small` data table, to help acquaint you with the variables that are included.
```{r}
DT::datatable(sample_n(Scorecard_small, size = 5))
```


In the code block below, examine the `Scorecard_small` data table to find out how many cases there are.

```{r one, exercise = TRUE}
nrow(Scorecard_small)
```

Adjust the axes, remove the "other" control, set the alpha so that the points don't overlie one another.

- look at `median_debt ~ net_price + color:control`
- look at `net_price_q1 ~ net_price + color:control`
- look at `net_price_q1 ~ net_price_q4 + color:control`

```{r}
Scorecard_small %>%
  filter(control %in% c("other", "public", "private")) %>%
  #gf_point(median_debt ~ net_price + color:control + alpha:1) %>%
  gf_point(net_price_q1 ~ net_price + color:control + alpha:0.1)
  gf_lims(x = c(0, 40000))
```

QUESTION: Expensive private colleges often claim that, because of financial aid, it can be less expensive to attend them than public schools. 
- Is there evidence for this in terms of the net price for low-income students compared to institutional average net price?
- Compare q1 to q5 net prices. Which private schools have q1 comparable to public institutions.


For later:

* Join in the Carnegie codes. Look at median debt versus net_price using color for public private and facet for large, small, ...
